This document imports, processes, and merges the SocioBosque CSV files for the 900- and 300-point samples.

**NA convention used throughout this pipeline and all downstream analysis:**
missing / not-applicable is always the real R `NA` — never the literal text `"NA"`.
Categorical columns (`sociobosque_type`, `pa`, `indigenous`, `within_sociobosque`)
use real `NA` for "not treated" / "not in a protected area" / "not indigenous land" /
"unknown". Downstream code should test with `is.na(x)` / `!is.na(x)`, never
`x == "NA"` — string comparisons against `"NA"` are fragile and were the source of
earlier bugs. CSVs are written with `na = ""` (blank cells) so no ambiguous `"NA"`
text is ever persisted to disk.

```{r setup, message=FALSE, warning=FALSE}
library(readr)
library(dplyr)

project_dir <- "/Users/chrissiepantoja/Library/CloudStorage/GoogleDrive-chrissiepantoja@gmail.com/My Drive/SocioBosques_Ecuador"
csv_dir <- file.path(project_dir, "processed_data", "csvs")

amazon_provinces <- c(
  "MORONA SANTIAGO", "NAPO", "ORELLANA", "PASTAZA",
  "SUCUMBIOS", "ZAMORA CHINCHIPE"
)
```

## Protected-area years

```{r protected-area-years}
pa_years <- c(
  "ANTISANA" = 1993, "CAJAS" = 1997, "CAYAMBE COCA" = 1970,
  "CERRO PLATEADO" = 2010, "CHIMBORAZO" = 1987,
  "COFAN BERMEJO" = 2002, "COLONSO CHALUPAS" = 2014,
  "CORDILLERA ORIENTAL DEL CARCHI" = 2019, "COTOPAXI" = 1975,
  "CUYABENO" = 1979, "EL BOLICHE" = 1979, "EL CONDOR" = 1999,
  "EL QUIMI" = 2006, "EL ZARZA" = 2006, "LA BONITA" = 2017,
  "LIMONCOCHA" = 1985, "LLANGANATES" = 1996,
  "LOS ILINIZAS" = 1996, "MARCOS PEREZ DE CASTILLA" = 1999,
  "PODOCARPUS" = 1982, "QUIMSACOCHA" = 2012,
  "RIO NEGRO SOPLADORA" = 2018, "SANGAY" = 1975,
  "SIETE IGLESIAS" = 2012, "SUMACO NAPO-GALERAS" = 1994,
  "TAMBILLO" = 2018, "YACURI" = 2009, "YASUNI" = 1979
)
```

## Helper functions

```{r helper-functions}
# na = c("", "NA") is explicit here (it's readr's default too) so that any
# literal "NA" text or blank cell in the raw source files becomes a real R NA
# right at ingest, before it can propagate through merges and comparisons.
#
# guess_max = Inf forces read_csv to scan the ENTIRE file (not just the first
# 1000 rows) when guessing each column's type. With guess_max left at its
# default, a column type guessed from an early homogeneous slice of the file
# (e.g. if SocioBosque-enrolled points happen to be listed first) can cause
# every value elsewhere that doesn't fit that guess to be silently turned into
# NA or garbled — exactly the "one or more parsing issues" warning, and the
# likely cause of sociobosque_type/within_sociobosque looking artificially
# uniform. If a file is very large this is slower, but correctness comes first
# here; problems() is also called immediately so any remaining issue is
# visible right away instead of silently downstream.
read_csv_file <- function(name) {
  path <- file.path(csv_dir, name)
  dat <- read_csv(path, na = c("", "NA"), guess_max = Inf, show_col_types = FALSE)
  probs <- problems(dat)
  if (nrow(probs) > 0) {
    warning(
      "Parsing problems in ", name, " (", nrow(probs), " issue(s)). ",
      "First few:\n",
      paste(capture.output(print(head(probs, 5))), collapse = "\n")
    )
  }
  dat
}

matched_merge <- function(x, y) {
  if (anyDuplicated(x$pointid) || anyDuplicated(y$pointid)) {
    stop("pointid must be unique in both datasets for a 1:1 merge.")
  }
  inner_join(x, y, by = "pointid")
}

encode_integer <- function(x) {
  as.integer(factor(x))
}
```

## Process forest data

```{r process-forest}
process_forest <- function(size) {
  prefix <- paste0("sample_", size, "_")
  forest <- read_csv_file(paste0(prefix, "forest.csv"))

  for (year in 1997:2020) {
    previous_year <- year - 1

    forest[[paste0("forestloss", year)]] <-
      forest[[paste0("mapbiomas", previous_year)]] -
      forest[[paste0("mapbiomas", year)]]

    loss_name <- paste0("mapbiomasloss", year)
    if (loss_name %in% names(forest)) forest[[loss_name]] <- NULL

    forest[[paste0("relforestloss", year)]] <-
      forest[[paste0("forestloss", year)]] /
      forest[[paste0("mapbiomas", previous_year)]]
  }

  names(forest)[match(paste0("mapbiomas", 1996:2020), names(forest))] <-
    paste0("forestcover", 1996:2020)

  forest
}
```

## Process and merge one sample

```{r process-sample}
process_sample <- function(size) {
  prefix <- paste0("sample_", size, "_")
  data <- matched_merge(
    read_csv_file(paste0(prefix, "spatial_attr.csv")),
    process_forest(size)
  )

  farming <- read_csv_file(paste0(prefix, "outcome14.csv")) %>%
    select(-matches("^mapbiomasloss"))
  names(farming)[match(paste0("mapbiomas", 1996:2020), names(farming))] <-
    paste0("farming", 1996:2020)
  data <- matched_merge(data, farming)

  mining <- read_csv_file(paste0(prefix, "outcome24.csv")) %>%
    select(-matches("^mapbiomasloss"))
  names(mining)[match(paste0("mapbiomas", 1996:2020), names(mining))] <-
    paste0("mining", 1996:2020)
  data <- matched_merge(data, mining)

  infrastructure <- read_csv_file(paste0(prefix, "outcome30.csv")) %>%
    select(-matches("^mapbiomasloss"))
  names(infrastructure)[match(paste0("mapbiomas", 1996:2020), names(infrastructure))] <-
    paste0("infrastruct", 1996:2020)
  data <- matched_merge(data, infrastructure)

  covariates <- read_csv_file(paste0(prefix, "covs.csv")) %>%
    rename(village = poblados)
  data <- matched_merge(data, covariates)

  protected <- read_csv_file(paste0(prefix, "PA.csv")) %>%
    mutate(
      year_pa = unname(pa_years[nam]),
      protected_areas = encode_integer(subap),
      pa = subap
    ) %>%
    select(pointid, subap, year_pa, protected_areas, pa)
  data <- matched_merge(data, protected)

  indigenous <- read_csv_file(paste0(prefix, "indigenous_new.csv")) %>%
    select(pointid, status) %>%
    mutate(
      indigenous_land = encode_integer(status),
      indigenous = status,
      within_indigenous = as.integer(!is.na(status))
    )
  data <- matched_merge(data, indigenous)

  soil <- read_csv_file(paste0(prefix, "soil.csv"))

  if (!"shape_area" %in% names(soil)) {
    possible_area_names <- c(
      "shape_area",
      "shapearea",
      "area",
      "area_m2",
      "shape_area_1",
      "area_ha"
    )

    area_name <- names(soil)[
      match(tolower(possible_area_names), tolower(names(soil)))
    ]
    area_name <- area_name[!is.na(area_name)][1]

    if (is.na(area_name)) {
      stop(
        "The soil file does not contain a shape-area column.\n",
        "Available columns are: ",
        paste(names(soil), collapse = ", ")
      )
    }

    soil <- soil %>%
      rename(shape_area = all_of(area_name))
  }

  soil <- soil %>%
    select(pointid, pen_cod, text1_cod, shape_area) %>%
    mutate(
      slope = encode_integer(pen_cod),
      soil = encode_integer(text1_cod),
      polygonid = encode_integer(as.character(shape_area))
    ) %>%
    select(-pen_cod, -text1_cod, -shape_area)
  data <- matched_merge(data, soil)

  canton_raw <- read_csv_file(paste0(prefix, "cantones.csv"))

  canton_id_candidates <- c("dpa_canton", "dpa_can", "canton_id")
  canton_name_candidates <- c("dpa_descan", "dpa_des", "canton", "canton_name")

  canton_id_name <- names(canton_raw)[
    match(tolower(canton_id_candidates), tolower(names(canton_raw)))
  ]
  canton_id_name <- canton_id_name[!is.na(canton_id_name)][1]

  canton_desc_name <- names(canton_raw)[
    match(tolower(canton_name_candidates), tolower(names(canton_raw)))
  ]
  canton_desc_name <- canton_desc_name[!is.na(canton_desc_name)][1]

  if (is.na(canton_id_name) || is.na(canton_desc_name)) {
    stop(
      "The cantones file does not contain the expected canton id/name columns.\n",
      "Available columns are: ",
      paste(names(canton_raw), collapse = ", ")
    )
  }

  canton <- canton_raw %>%
    select(pointid, all_of(canton_id_name), all_of(canton_desc_name)) %>%
    rename(canton = all_of(canton_desc_name), cantonid = all_of(canton_id_name))
  data <- matched_merge(data, canton)

  coordinates <- read_csv_file(paste0(prefix, "coord.csv"))

  data %>%
    matched_merge(coordinates) %>%
    mutate(
      state = toupper(trimws(as.character(state)))
    ) %>%
    filter(state %in% amazon_provinces) %>%
    mutate(
      stateid = encode_integer(state),
      sbp_type = encode_integer(sociobosque_type),
      # within_sociobosque arrives as text ("TRUE"/"FALSE"/blank/"NA") from the
      # raw files; coerce to a real logical so downstream code can use it
      # directly (if (within_sociobosque)) or test missingness with is.na(),
      # instead of comparing against the string "TRUE"/"NA".
      within_sociobosque = case_when(
        toupper(trimws(as.character(within_sociobosque))) == "TRUE" ~ TRUE,
        toupper(trimws(as.character(within_sociobosque))) == "FALSE" ~ FALSE,
        TRUE ~ NA
      ),
      withinsb = as.integer(within_sociobosque),
      sociobosque_year = parse_integer(as.character(sociobosque_year)),
      cantonid = parse_integer(as.character(cantonid)),
      elevation = parse_number(as.character(elevation)),
      # explicit, unambiguous indicator flags — TRUE/FALSE, never a string to compare against
      has_pa = !is.na(pa),
      has_indigenous = !is.na(indigenous)
    )
}
```

## Create the samples

```{r create-samples}
SBP_wide_900 <- process_sample(900)
SBP_wide_300 <- process_sample(300)
```

## Export to CSV

```{r export-csv}
# na = "" writes missing values as blank cells, never as the text "NA" —
# this is what makes "NA" unambiguous downstream (it never appears as data).
write_csv(
  SBP_wide_900,
  file.path(project_dir, "SBP_wide_900.csv"),
  na = ""
)

write_csv(
  SBP_wide_300,
  file.path(project_dir, "SBP_wide_300.csv"),
  na = ""
)
```

## Check the merged datasets

```{r check-results}
glimpse(SBP_wide_900)
glimpse(SBP_wide_300)

nrow(SBP_wide_900)
nrow(SBP_wide_300)
```



