---
title: "Mahalanobis Distance Matching & Covariate Balance Diagnostics"
author: "Andrew (Daye) Zhai & Chrissie A. Pantoja Vallejos"
subtitle: "Sociobosque, Ecuador"
output: html_document
---

This R Markdown reproduces the Stata do-file's Mahalanobis distance matching (MDM)
pipeline. Stata's `psmatch2, mahalanobis(...)` is replaced with `MatchIt::matchit()`
using `distance = "mahalanobis"`, `method = "nearest"`, `ratio = 1`, and a caliper —
the closest R equivalent to `n(1) caliper(0.1)`.

Rather than copy-pasting the same matching block once per subgroup (as the .do file
does ~10 times for Collective / Individual / PA populations, each split into
SB-only / SB+IT / SB+PA / SB+IT+PA), the logic is written once as a set of
functions and then called with different filters. This section fully implements
**300×300, Collective, Never-treated** (all four sub-restrictions).

```{r setup, message=FALSE, warning=FALSE}
library(readr)
library(dplyr)
library(tidyr)
library(purrr)
library(MatchIt)

project_dir <- "/Users/chrissiepantoja/Library/CloudStorage/GoogleDrive-chrissiepantoja@gmail.com/My Drive/SocioBosques_Ecuador"
data_dir <- project_dir

covariates <- c("road", "village", "river", "population",
                "oil", "pipelines", "slope", "elevation")
cv <- c("farming", "mining", "infrastruct")
```

## Core matching functions

### `mdm_cohort_never()` — matches never-treated vs. cohort-treated, within state

Translates the `mdm_cohort_never` Stata program. Adds up to 10 years of
pre-treatment `forestloss` covariates (any that exist for `cohort-10` to
`cohort-1`), builds treated/control flags, then matches separately within each
state.

```{r mdm-cohort-never}
mdm_cohort_never <- function(data, cohort, covariates) {

  pre_years <- (cohort - 10):(cohort - 1)
  loss_vars <- paste0("forestloss", pre_years)
  loss_vars <- loss_vars[loss_vars %in% names(data)]
  covars <- c(covariates, loss_vars)

  missing_covars <- covars[!covars %in% names(data)]
  if (length(missing_covars) > 0) {
    message("Missing covariates: ", paste(missing_covars, collapse = ", "))
    return(NULL)
  }

  d <- data %>%
    mutate(
      treated = as.integer(first_treat == cohort &
                              within_sociobosque == TRUE &
                              pixel_percent_intersect_sociobosque >= 0.5),
      control = as.integer(first_treat == 0 |
                              is.na(within_sociobosque) |
                              pixel_percent_intersect_sociobosque < 0.5)
    ) %>%
    filter(treated == 1 | control == 1)

  message("\n========== Processing cohort ", cohort, " ==========")
  n_treated <- sum(d$treated == 1, na.rm = TRUE)
  message("Treated group samples: ", n_treated)
  if (n_treated <= 1) {
    message("Insufficient treated units for cohort ", cohort, ". Skipping.")
    return(NULL)
  }
  n_control <- sum(d$control == 1, na.rm = TRUE)
  message("Control group candidates: ", n_control)
  if (n_control == 0) {
    message("No control candidates for cohort ", cohort, ". Skipping.")
    return(NULL)
  }

  states <- d %>%
    filter(treated == 1, !is.na(state)) %>%
    distinct(state) %>%
    pull(state)

  matched_list <- map(states, function(s) {
    message("========== Matching within cohort ", cohort, ", state = ", s, " ==========")
    ds <- d %>% filter(state == s)

    n_t <- sum(ds$treated == 1, na.rm = TRUE)
    message("Treated group samples: ", n_t)
    if (n_t <= 1) {
      message("Insufficient treated pixels in state = ", s, ". Skipping.")
      return(NULL)
    }
    n_c <- sum(ds$control == 1, na.rm = TRUE)
    message("Control group candidates: ", n_c)
    if (n_c <= 2) {
      message("Insufficient control group candidates in state = ", s, ". Skipping.")
      return(NULL)
    }

    fml <- as.formula(paste("treated ~", paste(covars, collapse = " + ")))
    m <- tryCatch(
      matchit(fml, data = ds, distance = "mahalanobis",
              method = "nearest", ratio = 1, caliper = 0.1),
      error = function(e) {
        message("Matching failed for state = ", s, ": ", conditionMessage(e))
        NULL
      }
    )
    if (is.null(m)) return(NULL)

    md <- match.data(m)  # adds distance, weights, subclass
    md %>%
      select(pointid, x, y, polygonid, treated,
             all_of(covars), weights) %>%
      rename(`_treated` = treated, `_weight` = weights) %>%
      mutate(cohort_year = cohort, state_match = s)
  })

  all_matched <- bind_rows(matched_list)

  if (nrow(all_matched) > 0) {
    write_csv(all_matched, file.path(data_dir, paste0("matched_cohort_", cohort, ".csv")))
  } else {
    message("No matched pairs produced for cohort ", cohort, ".")
  }

  all_matched
}
```

### `mdm_cohort_never_pa()` — same, but also split by `year_pa` within state

Translates `mdm_cohort_never_pa`: identical to the function above, except
matching is nested by state *and* protected-area designation year.

```{r mdm-cohort-never-pa}
mdm_cohort_never_pa <- function(data, cohort, covariates) {

  pre_years <- (cohort - 10):(cohort - 1)
  loss_vars <- paste0("forestloss", pre_years)
  loss_vars <- loss_vars[loss_vars %in% names(data)]
  covars <- c(covariates, loss_vars)

  missing_covars <- covars[!covars %in% names(data)]
  if (length(missing_covars) > 0) {
    message("Missing covariates: ", paste(missing_covars, collapse = ", "))
    return(NULL)
  }

  d <- data %>%
    mutate(
      treated = as.integer(first_treat == cohort &
                              within_sociobosque == TRUE &
                              pixel_percent_intersect_sociobosque >= 0.5),
      control = as.integer(first_treat == 0 |
                              is.na(within_sociobosque) |
                              pixel_percent_intersect_sociobosque < 0.5)
    ) %>%
    filter(treated == 1 | control == 1)

  message("\n========== Processing cohort ", cohort, " ==========")
  n_treated <- sum(d$treated == 1, na.rm = TRUE)
  message("Treated group samples: ", n_treated)
  if (n_treated <= 1) {
    message("Insufficient treated units for cohort ", cohort, ". Skipping.")
    return(NULL)
  }
  n_control <- sum(d$control == 1, na.rm = TRUE)
  message("Control group candidates: ", n_control)
  if (n_control == 0) {
    message("No control group candidates for cohort ", cohort, ". Skipping.")
    return(NULL)
  }

  states <- d %>%
    filter(treated == 1, !is.na(state)) %>%
    distinct(state) %>%
    pull(state)

  matched_list <- map(states, function(s) {
    years_pa <- d %>%
      filter(treated == 1, state == s, !is.na(year_pa)) %>%
      distinct(year_pa) %>%
      pull(year_pa)

    map(years_pa, function(ypa) {
      message("========== Matching within cohort ", cohort, ", state = ", s,
              ", year_pa = ", ypa, " ==========")
      ds <- d %>% filter(state == s, year_pa == ypa)

      n_t <- sum(ds$treated == 1, na.rm = TRUE)
      message("Treated group samples: ", n_t)
      if (n_t <= 1) {
        message("Insufficient treated pixels in state = ", s,
                " and year_pa = ", ypa, ". Skipping.")
        return(NULL)
      }
      n_c <- sum(ds$control == 1, na.rm = TRUE)
      message("Control group candidates: ", n_c)
      if (n_c <= 2) {
        message("Insufficient control group candidates in state = ", s,
                " and year_pa = ", ypa, ". Skipping.")
        return(NULL)
      }

      fml <- as.formula(paste("treated ~", paste(covars, collapse = " + ")))
      m <- tryCatch(
        matchit(fml, data = ds, distance = "mahalanobis",
                method = "nearest", ratio = 1, caliper = 0.1),
        error = function(e) {
          message("Matching failed for state = ", s, ", year_pa = ", ypa, ": ",
                  conditionMessage(e))
          NULL
        }
      )
      if (is.null(m)) return(NULL)

      md <- match.data(m)
      md %>%
        select(pointid, x, y, polygonid, treated, state, year_pa,
               all_of(covars), weights) %>%
        rename(`_treated` = treated, `_weight` = weights) %>%
        mutate(cohort_year = cohort, state_match = s, pa_match_year = ypa)
    })
  }) %>% flatten()

  all_matched <- bind_rows(matched_list)

  if (nrow(all_matched) > 0) {
    write_csv(all_matched, file.path(data_dir, paste0("matched_cohort_", cohort, ".csv")))
  } else {
    message("No matched pairs produced for cohort ", cohort, ".")
  }

  all_matched
}
```

## Generalized pipeline (replaces the repeated copy-pasted blocks)

`run_mdm_pipeline()` reproduces steps 2–6 that appear after every `di "*==...=="`
header in the .do file: run matching per cohort, consolidate matched pairs, drop
unmatched observations, reshape to long, save. `restriction_filter` is a
tidy-eval expression standing in for lines like `keep if pa == "NA" & indigenous == "NA"`.

```{r run-pipeline}
run_mdm_pipeline <- function(data, years, restriction_filter, mdm_fun,
                              covariates, base_data_for_reshape, label) {

  # Step 2: matching per cohort
  matched_by_cohort <- map(years, function(yr) {
    d <- data %>% filter(!!restriction_filter)
    message("========== Processing cohort ", yr, " ==========")
    mdm_fun(d, yr, covariates)
  })

  # Step 3: consolidate matched pairs
  matched_by_cohort_nonempty <- compact(matched_by_cohort)
  if (length(matched_by_cohort_nonempty) == 0) {
    warning("No cohort produced any matched units for '", label,
            "' — every cohort had 0 treated or 0 control candidates. ",
            "See the per-cohort messages above. Returning NULL for this label.")
    return(NULL)
  }

  all_matched_raw <- bind_rows(matched_by_cohort_nonempty) %>%
    distinct(pointid, cohort_year, .keep_all = TRUE) %>%
    arrange(pointid, cohort_year, `_treated`) %>%
    filter(!is.na(`_treated`))  # rule out obs not included in matching

  write_csv(all_matched_raw,
            file.path(data_dir, paste0("all_unmatched_pairs_wide_300_", label, ".csv")))

  # Step 4: keep only common-support / weighted matches
  all_matched <- all_matched_raw %>% filter(!is.na(`_weight`))
  write_csv(all_matched,
            file.path(data_dir, paste0("all_matched_pairs_wide_300_", label, ".csv")))

  # Step 5: clean temp per-cohort files
  walk(years, function(yr) {
    f <- file.path(data_dir, paste0("matched_cohort_", yr, ".csv"))
    if (file.exists(f)) file.remove(f)
  })

  # Step 6: reshape wide -> long and merge back full covariate history
  long_data <- all_matched %>%
    select(-starts_with("forestloss")) %>%
    mutate(newid = paste(pointid, cohort_year, sep = "_")) %>%
    inner_join(base_data_for_reshape, by = "pointid", suffix = c("", ".full")) %>%
    pivot_longer(
      cols = matches("^(forestcover|forestloss|relforestloss|farming|mining|infrastruct)\\d{4}$"),
      names_to = c(".value", "year"),
      names_pattern = "^(forestcover|forestloss|relforestloss|farming|mining|infrastruct)(\\d{4})$"
    ) %>%
    mutate(year = as.integer(year),
           panel_id = as.integer(factor(newid)))

  write_csv(long_data, file.path(data_dir, paste0("SBP_long_300_did_", label, ".csv")))
  message("Completed: ", label)
  long_data
}
```

## 300×300, Collective, Never-treated

```{r collective-setup}
sbp_wide_300 <- read_csv(file.path(data_dir, "SBP_wide_300.csv"), show_col_types = FALSE)

cover_vars <- paste0("forestcover", 1997:2020)

sbp_wide_300_collective <- sbp_wide_300 %>%
  mutate(
    treatment_year = if_else(is.na(sociobosque_year), 9999L, sociobosque_year),
    first_treat = if_else(is.na(sociobosque_year), 0L, sociobosque_year)
  ) %>%
  filter(sociobosque_type == "COLECTIVO" | is.na(sociobosque_type)) %>%
  mutate(max_cover_pre = pmax(!!!syms(cover_vars), na.rm = TRUE)) %>%
  filter(max_cover_pre > 0.3) %>%
  select(-max_cover_pre) %>%
  filter(!((is.na(within_sociobosque) & pixel_percent_intersect_sociobosque >= 0.5) |
             (within_sociobosque == TRUE & pixel_percent_intersect_sociobosque < 0.5)))

write_csv(sbp_wide_300_collective, file.path(data_dir, "SBP_wide_300_Collective.csv"))

years <- sbp_wide_300_collective %>%
  filter(!is.na(sociobosque_year)) %>%
  distinct(sociobosque_year) %>%
  pull(sociobosque_year) %>%
  sort()
```

### SB Only

```{r collective-sb-only}
run_mdm_pipeline(
  data = sbp_wide_300_collective,
  years = years,
  restriction_filter = quo(is.na(pa) & is.na(indigenous)),
  mdm_fun = mdm_cohort_never,
  covariates = covariates,
  base_data_for_reshape = sbp_wide_300_collective,
  label = "Collective_mdm_Never-treated_sb"
)
```

### SB + IT

```{r collective-sb-it}
run_mdm_pipeline(
  data = sbp_wide_300_collective,
  years = years,
  restriction_filter = quo(!is.na(indigenous) & is.na(pa)),
  mdm_fun = mdm_cohort_never,
  covariates = covariates,
  base_data_for_reshape = sbp_wide_300_collective,
  label = "Collective_mdm_Never-treated_sb+it"
)
```

### SB + PA

```{r collective-sb-pa}
run_mdm_pipeline(
  data = sbp_wide_300_collective,
  years = years,
  restriction_filter = quo(!is.na(pa) & is.na(indigenous)),
  mdm_fun = mdm_cohort_never_pa,
  covariates = covariates,
  base_data_for_reshape = sbp_wide_300_collective,
  label = "Collective_mdm_Never-treated_sb+pa"
)
```

### SB + IT + PA

```{r collective-sb-it-pa}
run_mdm_pipeline(
  data = sbp_wide_300_collective,
  years = years,
  restriction_filter = quo(!is.na(pa) & !is.na(indigenous)),
  mdm_fun = mdm_cohort_never_pa,
  covariates = covariates,
  base_data_for_reshape = sbp_wide_300_collective,
  label = "Collective_mdm_Never-treated_sb+it+pa"
)
```

### Covariate balance diagnostics for the whole sample

The `pstest`/`asdoc` Word export and `graph combine` figure are Stata-specific
reporting steps. The R equivalent is the `cobalt` package's `bal.tab()` /
`love.plot()`, working off the four appended matched files:

```{r covariate-balance, eval=FALSE}
library(cobalt)

all_collective <- bind_rows(
  read_csv(file.path(data_dir, "all_unmatched_pairs_wide_300_Collective_mdm_Never-treated_sb.csv"), na = "", show_col_types = FALSE),
  read_csv(file.path(data_dir, "all_unmatched_pairs_wide_300_Collective_mdm_Never-treated_sb+it.csv"), na = "", show_col_types = FALSE),
  read_csv(file.path(data_dir, "all_unmatched_pairs_wide_300_Collective_mdm_Never-treated_sb+pa.csv"), na = "", show_col_types = FALSE),
  read_csv(file.path(data_dir, "all_unmatched_pairs_wide_300_Collective_mdm_Never-treated_sb+it+pa.csv"), na = "", show_col_types = FALSE)
)

# love.plot / bal.tab replace pstest's dot chart, histogram, and scatter plot
bal.tab(`_treated` ~ ., data = all_collective, weights = all_collective$`_weight`,
        method = "matching")
```

## Not yet converted

This file is one section of a much larger (~1,600-line) pipeline. Still to
translate, if useful:

- **Individual, Never-treated** and **PA, Never-treated** sections (lines
  507–928) — identical pattern to above, just swap the `sociobosque_type`/`pa`
  filter, so they can reuse `run_mdm_pipeline()` directly.
- **`mdm_cohort_itcc_sbit`** and **`mdm_cohort_itcc_sbit_state`** programs
  (Individual-as-treated, Collective-as-control) — a third matching variant,
  lines 1005–1156.
- The **Mata block** (lines 453–478) that repacks each row's forestloss values
  into `forestloss_pre1`–`forestloss_pre10` — straightforward in R with
  `apply()`/`pmap()`, but I didn't want to guess at exact behavior without
  seeing it used downstream.
- **`dcdh_effect_export_rate`** (end of file) — a post-estimation helper for a
  De Chaisemartin–D'Haultfœuille DiD estimator; this would map to whichever R
  package you're using for that estimator (e.g. `DIDmultiplegt`), not `MatchIt`.

Let me know which of these you want next, or if you'd like the whole file
finished in one pass.
